Mixing time estimation in reversible Markov chains from a single sample path

### 2015-06-02 upperbound.tex
ln -> log

% \ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}

Hence, with probability $1-\delta$, we have for all $i\in [d]$,
%\todoc{Is it possible to have $\pi_i$ here instead of $\pimin$ in the RHS? Probably not..}
\begin{align}
\label{eq:singletonboundsimple}
\Abs{\hat{\pi}_i-\pi_i}
\le
\sqrt{ 8 \pi_i(1-\pi_i) 
\,\frac{
\ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
  }{\gap n}}
+
\frac{20
  \ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
}{\gap n}
%\frac1{(n-1)\gap\sqrt\pimin}
%+
%\sqrt{\frac2{\pimin\gap^2n}\ln(2d/\delta)}
\le
23 \sqrt{  \pi_i 
	\frac{\ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
           }{\gap n}
       }\,,
\end{align}
where the second inequality holds as soon as 
\[
n \ge
	\frac{\ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
           }{\gap \pimin}\,.
\]
Consider now the following lemma:
\begin{lemma}
Let  $\pi_i,\wh\pi_i$, $i\in [2]$, $C>0$ be
such that $|\pi_i - \wh\pi_i| \le C \sqrt{\pi_i}$, $i\in [2]$.
Assume that $\pi_1\le \pi_2$.
Then, if $C\le 2\sqrt{\pi_1}$ then $\Abs{\pi_1 - \min(\wh\pi_1,\wh\pi_2) } \le C \sqrt{\pi_1}$.
\end{lemma}
\begin{proof}
Let $\hatpimin = \min(\wh\pi_1,\wh\pi_2)$.
We have
\begin{align*}
\hatpimin \le \hat{\pi}_1 \le  \pi_1 + C \sqrt{\pi_1}
\end{align*}
so it remains to show that
\begin{align*}
\pi_1 - C \sqrt{\pi_1} \le \hatpimin\,.
\end{align*}
Note that $\hat\pi_2 \ge \pi_2- C \sqrt{\pi_2 } $. Now,
our condition on $C$ implies that $C \le 2 \sqrt{\pi_1} = \sqrt{\pi_1}+\sqrt{\pi_2} = \frac{\pi_2 - \pi_1}{\sqrt{\pi_2}-\sqrt{\pi_1}}$. 
\begin{align*}
\pi_2 - C \sqrt{\pi_2  } \ge \pi_1 - C \sqrt{\pi_1 }
\end{align*}
holds if and only if
%\begin{align*}
%\pi_i - \pi_1 \ge C (\sqrt{\pi_i} - \sqrt{\pi_1}) \sqrt{\eps}\,,
%\end{align*}
%which in turn is equivalent to 
\begin{align*}
\sqrt{\pi_2} + \sqrt{\pi_1} = \frac{\pi_2 - \pi_1}{\sqrt{\pi_2} -\sqrt{\pi_1}} \ge C \,,
\end{align*}
which is implied by our condition on $\eps$. Thus, under our condition $\hat\pi_2 \ge \pi_2 - C  \sqrt{\pi_2 } \ge \pi_1 - C \sqrt{\pi_1}$, finishing the proof.
\end{proof}
By this lemma, as soon as $n\ge (11.5)^2 \frac{\ln( \frac{d}{\delta} )+\frac12\ln(\frac2\pimin) }{ \gap\pimin}$,
then
\[
\Abs{\pimin - \hatpimin} \le 23 \sqrt{ \pimin \frac{\ln( \frac{d}{\delta} )+\frac12\ln(\frac2\pimin) }{ \gap n} }\,,
\]
For $n$ smaller than the above critical value, the second term in \eqref{eq:piminbound} is lower bounded by one.
Noting that $\Abs{\pimin-\hatpimin}\le 1$ thus shows that \eqref{eq:piminbound} holds for any $n$.

@@

Let us now turn to proving \eqref{eq:gapbound}, i.e., 
the bound on the error of the spectral gap estimate $\hatgap$.
The accuracy of $\hatgap$ is based on the accuracy of $\Sym(\wh\vL)$
in approximating $\vL$ via Weyl's inequality:
\begin{align*}%\label{eq:Weyl}
  |\hat\lambda_i - \lambda_i|
  \leq \norm{\Sym(\wh\vL) - \vL}
  \quad \text{for all }i \in [d] .
\end{align*}
Moreover, the triangle inequality implies that symmetrizing $\wh\vL$
can only help:
\begin{align*}
%\label{eq:symm}
  \norm{\Sym(\wh\vL) - \vL} \leq \norm{\wh\vL - \vL} .
\end{align*}
Therefore, the bound on the spectral gap in \cref{thm:err} is readily obtained from the
following lemma which bounds $\norm{\wh\vL - \vL}$ as well as $\abs{
\hat\pi_i/\pi_i - 1 }$ for each $i \in [d]$ with high probability.
\begin{lemma}
  \label{lem:err}
  There exists an absolute constant $C>0$ such that the following
  holds.
  For any $\delta \in (0,1)$, with probability at least $1-\delta$,
  \begin{equation*}
    \norm{\wh\vL - \vL}
    \leq
    \frac{C}{\pimin}
      \sqrt{\frac{\ln(d/\delta)\ln\frac{n}{\delta\pimin}}{n\pimin\gap}} \,.
  \end{equation*}
	 \todoc{Note that the bound that was here essentially got multiplied by $1/\pimin$!}
\end{lemma}
The remainder of this section is devoted to proving this \lcnamecref{lem:err}.

@@

\subsection{A bound on $\norm{\errp}$}
We have
\begin{align*}
  \norm{\errp}
  = \max_{i\in[d]} \Abs{
    \sqrt{\frac{\pi_i}{\hat\pi_i}} - 1\,.
  }
\end{align*}
%The purpose of this section is to develop a bound on the RHS of \eqref{eq:singleton} so as to bound $\norm{\errp}$.
First note that for $x$ real, $\Abs{x-1}\le 1/2$ implies that $\max(\Abs{x-1},\Abs{1/x-1}) \le 2 \Abs{x-1}$.
Hence, it suffices to develop a bound on $ \Abs{ \frac{\hat\pi_i}{\pi_i} - 1 }$;
the final bound will follow under the condition that the developed bound is below $1/2$.
The required bound is readily obtained from~\eqref{eq:singletonboundsimple}:
As long as $n = \Omega \Parens{
	\frac{\ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
           }{\gap \pimin} }$, \todoc{First condition on $n$.}
$
\Abs{\frac{\hat{\pi}_i}{\pi_i}-1}
\le
23 \sqrt{  
	\frac{\ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
           }{\pimin\gap n}
       }\,,
$
and thus
\begin{align}
\label{eq:singletonbound}
\norm{\errp} \le C \sqrt{ \frac{\ln\Parens{\frac{d}\delta} + \ln \frac{1}{\pimin} }{\pimin\gap n}} \,.
\end{align}
We now turn to bounding  $\norm{\errm}$.
%Note that although the entries of $\wh{\vM}$ and $\hat\vpi$ are simply
%sample averages, they are formed using a \emph{single sample path}
%from a (possibly) non-stationary Markov chain.
%Applying Chernoff-Hoeffding bounds for Markov chains to each entry of
%$\wh{\vM}$ would result in a
%significantly worse sample complexity bound.
%%vastly worse sequence length requirement.
%Instead, we adapt probability tail bounds for sums of independent
%random matrices to our setting.

@@

Suppose the first block $F$ is of length
\begin{equation}
  \label{eq:block-length}
  |F| = a \geq 1 + \frac{1}{\gap} \ln \frac1{\pimin\gap}
  .
\end{equation}
Then
\[
  \frac{(1-\gap)^{\min H_s - 1}}{\pimin\gap}
  \leq 1
\]
for all blocks $H_s$ (and same for the $T_s$ blocks), so
\begin{align}
  \max \Braces{
    \Abs{ \frac{\pi_i^{(H_s)}}{\pi_i} } ,\,
    \Abs{ \frac{\pi_i^{(T_s)}}{\pi_i} } ,\,
    \Abs{ \frac{\pi_i^{(H,T)}}{\pi_i} }
    : i \in [d] , s
  }
  & \leq 2
  .
  \label{eq:near-stationary}
\end{align}
This means that as long as the blocks are sufficiently long, then all
but the first block $F$ can effectively be thought of as being started
from the stationary distribution.
We shall henceforth assume that~\eqref{eq:block-length} holds.
\fi
\subsubsection{
Bounding $\norm{\errm-\EE{\errm}}$: 
Application of a matrix tail inequality}
\label{sec:pairwise-tailbound}

@@

Let
\[
  \vY_s := \frac1{|H_s|} \sum_{t \in H_s} \ve_{X_t} \ve_{X_{t+1}}^\t ,
  \quad s \in [\mu_H] ,
\]
so
\[
  \wh{\vM}_H = \sum_{s=1}^{\mu_H} w_{H,s} \vY_s ,
\]
a weighted average of the random matrices $\vY_s$.
For each $s \in [\mu_H]$, the random matrix $\vY_s$ is a function of
\[ (X_t : (2s-1)a + 1 \leq t \leq 2sa + 1) \]
(note the $+1$ in the upper limit of $t$).
So $\vY_s$ and $\vY_{s+1}$ are separated by $a-1$ time steps.
When $a$ is sufficiently large, we will be able to effectively treat
the random matrices $\vY_s$ as if they were independent.
In the sequel, we shall always assume that the block length $a$
satisfies
\begin{equation}
  \label{eq:block-length}
  a \geq
   1 + \frac{1}{2\gap} \ln \frac1{\pimin}\,.
%   \geq
%   1 + \frac{1}{\gap} \ln \frac1{|H_s|\sqrt{\pimin}\gap}
\end{equation}
\todoc{By using the geometric series bound,
we can derive that it is enough
if $a\ge 1 + \frac{1}{\gap} \ln \frac1{|H_s|\sqrt{\pimin}\gap}$,
which is better when $|H_s|$ is bigger than $\gap$.
}
\todoc{As we need $a\le n/2$, we will need $n = \Omega(1/\gap\ln\frac1{\pimin})$.}

@@

Define
\[
  \vZ_s
  := \Diag(\vpi)^{-1/2} \left( \vY_s 
  - \bbE(\vY_s) \right)\Diag(\vpi)^{-1/2}
  .
\]
We apply a matrix tail inequality to the
$w_{H,s}$-weighted sum of \emph{independent} copies of the $\vZ_s$'s.
More precisely, we will apply the tail inequality to independent copies $\wt{\vZ}_s$, $s\in \iset{\mu_H}$ 
of the random variables $\vZ_s$ and then relate the weighted sum of $\wt{\vZ}_s$ to that of $\vZ_s$.
The matrix tail inequality we use is as follows:
\begin{theorem}[Matrix Bernstein inequality, Theorem~6.1.1 of \citet{tropp2015intro}]
\label{thm:mxbernstein}
Let $\vX_1,\dots,\vX_n$ be a sequence of independent, random matrices with common dimension $d_1\times d_2$.
Assume that $\EE{\vX_i} = 0$ and $\Norm{\vX_i}\le R$ for each $1\le i \le n$. Let $\vS = \sum_{i=1}^n \vX_i$ and let 
\begin{align*}
v = \max\left\{ \norm{\bbE \textstyle\sum_i \vX_i \vX_i^\top  }, 
						      \norm{\bbE \textstyle\sum_i \vX_i^\top \vX_i  }
			\right\}\,.
\end{align*}
Then, for all $t\ge 0$, 
\[
\bbP\left( \Norm{\vS} \ge t \right) \le (d_1 + d_2) \exp\left( -\frac{t^2/2}{v+R t/3}\right)\,.
\]
\end{theorem}
By inverting the bound, we get that for any $0\le \delta \le 1$, with probability $1-\delta$, 
\[
\norm{\vS} \le \sqrt{2 v \ln \frac{d_1+d_2}{\delta}} + \frac{2R}{3} \ln\frac{d_1+d_2}{\delta}\,.
\]

@@

Combining this with
$\norm{\Diag(\vpi)^{-1/2} \vM\Diag(\vpi)^{-1/2}} = \norm{\vL}\le 1$,
%\if0
%
%Since
%\[
%  \Diag(\vpi)^{-1/2} \bbE(\vY_s) \Diag(\vpi)^{-1/2}
%  =
%  \Diag(\vpi)^{-1/2}
%  \Diag(\vpi^{(H_s)})
%  \Diag(\vpi)^{-1/2}
%  \vL
%\]
%and
%\[
%  \norm{\vL} \leq 1
%  ,
%\]
%it follows that
%\[
%  \norm{\Diag(\vpi)^{-1/2} \bbE(\vY_s) \Diag(\vpi)^{-1/2}}
%  \leq \max_{i \in [d]} \frac{\pi_i^{(H_s)}}{\pi_i}\,.
%%  \leq 2
%\]
%
%Suppose the first block $F$ is of length
%\begin{equation}
%  \label{eq:block-length}
%  |F| \geq 1 + \frac{1}{\gap} \ln \frac1{\pimin\gap}
%  .
%\end{equation}
%Then
%\[
%  \frac{(1-\gap)^{\min H_s - 1}}{\pimin\gap}
%  \leq 1
%\]
%for all blocks $H_s$ (and same for the $T_s$ blocks), so
%\begin{align}
%  \max \Braces{
%    \Abs{ \frac{\pi_i^{(H_s)}}{\pi_i} } ,\,
%    \Abs{ \frac{\pi_i^{(T_s)}}{\pi_i} } ,\,
%    \Abs{ \frac{\pi_i^{(H,T)}}{\pi_i} }
%    : i \in [d] , s
%  }
%  & \leq 2
%  .
%  \label{eq:near-stationary}
%\end{align}
%
%
%
%(using the assumption that~\eqref{eq:block-length} holds).
%Therefore
%\fi
it follows that
\begin{align}
\label{eq:evysbound}
\norm{\Diag(\vpi)^{-1/2} \bbE(\vY_s) \Diag(\vpi)^{-1/2}} \le 2
\end{align}
which, together with~\eqref{eq:vysbound}, gives
\[
  \norm{\vZ_s} \leq \frac1{\pimin} + 2
%  \max_{i\in[d]} \frac{\pi_i^{(H_s)}}{\pi_i}
  .
\]

@@

\paragraph{Independent copies bound.}
Let $\wt\vZ_s$ for $s \in [\mu_H]$ be independent copies of
$\vZ_s$ for $s \in [\mu_H]$.
Applying \cref{thm:mxbernstein} to the weighted sum of these random variables, we  get that 
with probability at least $1-\delta$, the spectral norm of
$\sum_{s=1}^{\mu_H} w_{H,s} \wt\vZ_s$ is at most
\begin{align*}
\MoveEqLeft  \Norm{\sum_{s=1}^{\mu_H} w_{H,s} \wt\vZ_s}
  \leq
  \sqrt{
    4\Parens{ \max_s w_{H,s} } \Parens{
      d_{\vP}
      + 2
    }
    \ln(4d/\delta)
  }
  \\
& \qquad\qquad\qquad  +
  \frac23 \max_s w_{H,s}
  \Parens{
    \frac1{\pimin} + 2
  }
  \ln(4d/\delta)
\end{align*}
where
\begin{align*}
  d_{\vP}
  & := \max_{i \in [d]} \sum_{j=1}^d \frac{P_{i,j}}{\pi_j}
  \leq
  \frac{1}{\pimin}
  \,.
\end{align*}

For simplicity, we assume that all of the blocks have length exactly $a$.
Under this assumption, 
the above bound (also with looser constants) gives that
with probability at least $1-\delta$:
\begin{equation}
\label{eq:indbound}
  \Norm{\frac1{\mu_H} \sum_{s=1}^{\mu_H} \wt\vZ_s}
  \leq
  \sqrt{
    \frac{12\ln(4d/\delta)}{\mu_H\pimin}
  }
  +
  \frac{2\ln(4d/\delta)}{\mu_H\pimin}
  \,.
\end{equation}

@@

It is not hard to see that $\beta(\bbP)$ in turn is upper bounded by 
the $(a-1)$-th beta mixing coefficient of the Markov chain $\{X_t\}$ (because the first
index in $H_{s+1}$ and the last index in $H_s$ are at a distance of $(a-1)$).
Hence, 
%of a signed measure $\mu$ over a measurable space 
%$(\Omega,\mathcal{A})$, defined by 
%$\Norm{\mu}_{\tv} = \frac12(\mu^+(\Omega)+\mu^-(\Omega))$, 
%where $\mu^+$ and $\mu^-$ are the positive and negative parts of measure $\mu$.
by a well-known result 
\[
  \beta(\bbP)
  \leq
  \frac{1}{\pimin} \exp(-(a-1)\gap)
\]
\citep[see, for instance,][]{bradley05}. \todoc{Which theorem? I did not find this result there. Can anyone please refine this citation?}
It follows that if the block length $a$ satisfies
\begin{equation}
  \label{eq:block-length2}
  a \geq 1 + \frac{1}{\gap}\ln\frac{n}{\delta\pimin}
  ,
\end{equation}
then the failure probability increases from $\delta$ to at most
$2\delta$.
%Note that this condition implies \eqref{eq:block-length}.

>>>>

% \log\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}

Hence, with probability $1-\delta$, we have for all $i\in [d]$,
%\todoc{Is it possible to have $\pi_i$ here instead of $\pimin$ in the RHS? Probably not..}
\begin{align}
\label{eq:singletonboundsimple}
\Abs{\hat{\pi}_i-\pi_i}
\le
\sqrt{ 8 \pi_i(1-\pi_i) 
\,\frac{
\log\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
  }{\gap n}}
+
\frac{20
  \log\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
}{\gap n}
%\frac1{(n-1)\gap\sqrt\pimin}
%+
%\sqrt{\frac2{\pimin\gap^2n}\log(2d/\delta)}
\le
23 \sqrt{  \pi_i 
	\frac{\log\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
           }{\gap n}
       }\,,
\end{align}
where the second inequality holds as soon as 
\[
n \ge
	\frac{\log\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
           }{\gap \pimin}\,.
\]
Consider now the following lemma:
\begin{lemma}
Let  $\pi_i,\wh\pi_i$, $i\in [2]$, $C>0$ be
such that $|\pi_i - \wh\pi_i| \le C \sqrt{\pi_i}$, $i\in [2]$.
Assume that $\pi_1\le \pi_2$.
Then, if $C\le 2\sqrt{\pi_1}$ then $\Abs{\pi_1 - \min(\wh\pi_1,\wh\pi_2) } \le C \sqrt{\pi_1}$.
\end{lemma}
\begin{proof}
Let $\hatpimin = \min(\wh\pi_1,\wh\pi_2)$.
We have
\begin{align*}
\hatpimin \le \hat{\pi}_1 \le  \pi_1 + C \sqrt{\pi_1}
\end{align*}
so it remains to show that
\begin{align*}
\pi_1 - C \sqrt{\pi_1} \le \hatpimin\,.
\end{align*}
Note that $\hat\pi_2 \ge \pi_2- C \sqrt{\pi_2 } $. Now,
our condition on $C$ implies that $C \le 2 \sqrt{\pi_1} = \sqrt{\pi_1}+\sqrt{\pi_2} = \frac{\pi_2 - \pi_1}{\sqrt{\pi_2}-\sqrt{\pi_1}}$. 
\begin{align*}
\pi_2 - C \sqrt{\pi_2  } \ge \pi_1 - C \sqrt{\pi_1 }
\end{align*}
holds if and only if
%\begin{align*}
%\pi_i - \pi_1 \ge C (\sqrt{\pi_i} - \sqrt{\pi_1}) \sqrt{\eps}\,,
%\end{align*}
%which in turn is equivalent to 
\begin{align*}
\sqrt{\pi_2} + \sqrt{\pi_1} = \frac{\pi_2 - \pi_1}{\sqrt{\pi_2} -\sqrt{\pi_1}} \ge C \,,
\end{align*}
which is implied by our condition on $\eps$. Thus, under our condition $\hat\pi_2 \ge \pi_2 - C  \sqrt{\pi_2 } \ge \pi_1 - C \sqrt{\pi_1}$, finishing the proof.
\end{proof}
By this lemma, as soon as $n\ge (11.5)^2 \frac{\log( \frac{d}{\delta} )+\frac12\log(\frac2\pimin) }{ \gap\pimin}$,
then
\[
\Abs{\pimin - \hatpimin} \le 23 \sqrt{ \pimin \frac{\log( \frac{d}{\delta} )+\frac12\log(\frac2\pimin) }{ \gap n} }\,,
\]
For $n$ smaller than the above critical value, the second term in \eqref{eq:piminbound} is lower bounded by one.
Noting that $\Abs{\pimin-\hatpimin}\le 1$ thus shows that \eqref{eq:piminbound} holds for any $n$.

@@

Let us now turn to proving \eqref{eq:gapbound}, i.e., 
the bound on the error of the spectral gap estimate $\hatgap$.
The accuracy of $\hatgap$ is based on the accuracy of $\Sym(\wh\vL)$
in approximating $\vL$ via Weyl's inequality:
\begin{align*}%\label{eq:Weyl}
  |\hat\lambda_i - \lambda_i|
  \leq \norm{\Sym(\wh\vL) - \vL}
  \quad \text{for all }i \in [d] .
\end{align*}
Moreover, the triangle inequality implies that symmetrizing $\wh\vL$
can only help:
\begin{align*}
%\label{eq:symm}
  \norm{\Sym(\wh\vL) - \vL} \leq \norm{\wh\vL - \vL} .
\end{align*}
Therefore, the bound on the spectral gap in \cref{thm:err} is readily obtained from the
following lemma which bounds $\norm{\wh\vL - \vL}$ as well as $\abs{
\hat\pi_i/\pi_i - 1 }$ for each $i \in [d]$ with high probability.
\begin{lemma}
  \label{lem:err}
  There exists an absolute constant $C>0$ such that the following
  holds.
  For any $\delta \in (0,1)$, with probability at least $1-\delta$,
  \begin{equation*}
    \norm{\wh\vL - \vL}
    \leq
    \frac{C}{\pimin}
      \sqrt{\frac{\log(d/\delta)\log\frac{n}{\delta\pimin}}{n\pimin\gap}} \,.
  \end{equation*}
	 \todoc{Note that the bound that was here essentially got multiplied by $1/\pimin$!}
\end{lemma}
The remainder of this section is devoted to proving this \lcnamecref{lem:err}.

@@

\subsection{A bound on $\norm{\errp}$}
We have
\begin{align*}
  \norm{\errp}
  = \max_{i\in[d]} \Abs{
    \sqrt{\frac{\pi_i}{\hat\pi_i}} - 1\,.
  }
\end{align*}
%The purpose of this section is to develop a bound on the RHS of \eqref{eq:singleton} so as to bound $\norm{\errp}$.
First note that for $x$ real, $\Abs{x-1}\le 1/2$ implies that $\max(\Abs{x-1},\Abs{1/x-1}) \le 2 \Abs{x-1}$.
Hence, it suffices to develop a bound on $ \Abs{ \frac{\hat\pi_i}{\pi_i} - 1 }$;
the final bound will follow under the condition that the developed bound is below $1/2$.
The required bound is readily obtained from~\eqref{eq:singletonboundsimple}:
As long as $n = \Omega \Parens{
	\frac{\log\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
           }{\gap \pimin} }$, \todoc{First condition on $n$.}
$
\Abs{\frac{\hat{\pi}_i}{\pi_i}-1}
\le
23 \sqrt{  
	\frac{\log\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
           }{\pimin\gap n}
       }\,,
$
and thus
\begin{align}
\label{eq:singletonbound}
\norm{\errp} \le C \sqrt{ \frac{\log\Parens{\frac{d}\delta} + \log \frac{1}{\pimin} }{\pimin\gap n}} \,.
\end{align}
We now turn to bounding  $\norm{\errm}$.
%Note that although the entries of $\wh{\vM}$ and $\hat\vpi$ are simply
%sample averages, they are formed using a \emph{single sample path}
%from a (possibly) non-stationary Markov chain.
%Applying Chernoff-Hoeffding bounds for Markov chains to each entry of
%$\wh{\vM}$ would result in a
%significantly worse sample complexity bound.
%%vastly worse sequence length requirement.
%Instead, we adapt probability tail bounds for sums of independent
%random matrices to our setting.

@@

Suppose the first block $F$ is of length
\begin{equation}
  \label{eq:block-length}
  |F| = a \geq 1 + \frac{1}{\gap} \log \frac1{\pimin\gap}
  .
\end{equation}
Then
\[
  \frac{(1-\gap)^{\min H_s - 1}}{\pimin\gap}
  \leq 1
\]
for all blocks $H_s$ (and same for the $T_s$ blocks), so
\begin{align}
  \max \Braces{
    \Abs{ \frac{\pi_i^{(H_s)}}{\pi_i} } ,\,
    \Abs{ \frac{\pi_i^{(T_s)}}{\pi_i} } ,\,
    \Abs{ \frac{\pi_i^{(H,T)}}{\pi_i} }
    : i \in [d] , s
  }
  & \leq 2
  .
  \label{eq:near-stationary}
\end{align}
This means that as long as the blocks are sufficiently long, then all
but the first block $F$ can effectively be thought of as being started
from the stationary distribution.
We shall henceforth assume that~\eqref{eq:block-length} holds.
\fi
\subsubsection{
Bounding $\norm{\errm-\EE{\errm}}$: 
Application of a matrix tail inequality}
\label{sec:pairwise-tailbound}

@@

Let
\[
  \vY_s := \frac1{|H_s|} \sum_{t \in H_s} \ve_{X_t} \ve_{X_{t+1}}^\t ,
  \quad s \in [\mu_H] ,
\]
so
\[
  \wh{\vM}_H = \sum_{s=1}^{\mu_H} w_{H,s} \vY_s ,
\]
a weighted average of the random matrices $\vY_s$.
For each $s \in [\mu_H]$, the random matrix $\vY_s$ is a function of
\[ (X_t : (2s-1)a + 1 \leq t \leq 2sa + 1) \]
(note the $+1$ in the upper limit of $t$).
So $\vY_s$ and $\vY_{s+1}$ are separated by $a-1$ time steps.
When $a$ is sufficiently large, we will be able to effectively treat
the random matrices $\vY_s$ as if they were independent.
In the sequel, we shall always assume that the block length $a$
satisfies
\begin{equation}
  \label{eq:block-length}
  a \geq
   1 + \frac{1}{2\gap} \log \frac1{\pimin}\,.
%   \geq
%   1 + \frac{1}{\gap} \log \frac1{|H_s|\sqrt{\pimin}\gap}
\end{equation}
\todoc{By using the geometric series bound,
we can derive that it is enough
if $a\ge 1 + \frac{1}{\gap} \log \frac1{|H_s|\sqrt{\pimin}\gap}$,
which is better when $|H_s|$ is bigger than $\gap$.
}
\todoc{As we need $a\le n/2$, we will need $n = \Omega(1/\gap\log\frac1{\pimin})$.}

@@

Define
\[
  \vZ_s
  := \Diag(\vpi)^{-1/2} \left( \vY_s 
  - \bbE(\vY_s) \right)\Diag(\vpi)^{-1/2}
  .
\]
We apply a matrix tail inequality to the
$w_{H,s}$-weighted sum of \emph{independent} copies of the $\vZ_s$'s.
More precisely, we will apply the tail inequality to independent copies $\wt{\vZ}_s$, $s\in \iset{\mu_H}$ 
of the random variables $\vZ_s$ and then relate the weighted sum of $\wt{\vZ}_s$ to that of $\vZ_s$.
The matrix tail inequality we use is as follows:
\begin{theorem}[Matrix Bernstein inequality, Theorem~6.1.1 of \citet{tropp2015intro}]
\label{thm:mxbernstein}
Let $\vX_1,\dots,\vX_n$ be a sequence of independent, random matrices with common dimension $d_1\times d_2$.
Assume that $\EE{\vX_i} = 0$ and $\Norm{\vX_i}\le R$ for each $1\le i \le n$. Let $\vS = \sum_{i=1}^n \vX_i$ and let 
\begin{align*}
v = \max\left\{ \norm{\bbE \textstyle\sum_i \vX_i \vX_i^\top  }, 
						      \norm{\bbE \textstyle\sum_i \vX_i^\top \vX_i  }
			\right\}\,.
\end{align*}
Then, for all $t\ge 0$, 
\[
\bbP\left( \Norm{\vS} \ge t \right) \le (d_1 + d_2) \exp\left( -\frac{t^2/2}{v+R t/3}\right)\,.
\]
\end{theorem}
By inverting the bound, we get that for any $0\le \delta \le 1$, with probability $1-\delta$, 
\[
\norm{\vS} \le \sqrt{2 v \log \frac{d_1+d_2}{\delta}} + \frac{2R}{3} \log\frac{d_1+d_2}{\delta}\,.
\]

@@

Combining this with
$\norm{\Diag(\vpi)^{-1/2} \vM\Diag(\vpi)^{-1/2}} = \norm{\vL}\le 1$,
%\if0
%
%Since
%\[
%  \Diag(\vpi)^{-1/2} \bbE(\vY_s) \Diag(\vpi)^{-1/2}
%  =
%  \Diag(\vpi)^{-1/2}
%  \Diag(\vpi^{(H_s)})
%  \Diag(\vpi)^{-1/2}
%  \vL
%\]
%and
%\[
%  \norm{\vL} \leq 1
%  ,
%\]
%it follows that
%\[
%  \norm{\Diag(\vpi)^{-1/2} \bbE(\vY_s) \Diag(\vpi)^{-1/2}}
%  \leq \max_{i \in [d]} \frac{\pi_i^{(H_s)}}{\pi_i}\,.
%%  \leq 2
%\]
%
%Suppose the first block $F$ is of length
%\begin{equation}
%  \label{eq:block-length}
%  |F| \geq 1 + \frac{1}{\gap} \log \frac1{\pimin\gap}
%  .
%\end{equation}
%Then
%\[
%  \frac{(1-\gap)^{\min H_s - 1}}{\pimin\gap}
%  \leq 1
%\]
%for all blocks $H_s$ (and same for the $T_s$ blocks), so
%\begin{align}
%  \max \Braces{
%    \Abs{ \frac{\pi_i^{(H_s)}}{\pi_i} } ,\,
%    \Abs{ \frac{\pi_i^{(T_s)}}{\pi_i} } ,\,
%    \Abs{ \frac{\pi_i^{(H,T)}}{\pi_i} }
%    : i \in [d] , s
%  }
%  & \leq 2
%  .
%  \label{eq:near-stationary}
%\end{align}
%
%
%
%(using the assumption that~\eqref{eq:block-length} holds).
%Therefore
%\fi
it follows that
\begin{align}
\label{eq:evysbound}
\norm{\Diag(\vpi)^{-1/2} \bbE(\vY_s) \Diag(\vpi)^{-1/2}} \le 2
\end{align}
which, together with~\eqref{eq:vysbound}, gives
\[
  \norm{\vZ_s} \leq \frac1{\pimin} + 2
%  \max_{i\in[d]} \frac{\pi_i^{(H_s)}}{\pi_i}
  .
\]

@@

\paragraph{Independent copies bound.}
Let $\wt\vZ_s$ for $s \in [\mu_H]$ be independent copies of
$\vZ_s$ for $s \in [\mu_H]$.
Applying \cref{thm:mxbernstein} to the weighted sum of these random variables, we  get that 
with probability at least $1-\delta$, the spectral norm of
$\sum_{s=1}^{\mu_H} w_{H,s} \wt\vZ_s$ is at most
\begin{align*}
\MoveEqLeft  \Norm{\sum_{s=1}^{\mu_H} w_{H,s} \wt\vZ_s}
  \leq
  \sqrt{
    4\Parens{ \max_s w_{H,s} } \Parens{
      d_{\vP}
      + 2
    }
    \log(4d/\delta)
  }
  \\
& \qquad\qquad\qquad  +
  \frac23 \max_s w_{H,s}
  \Parens{
    \frac1{\pimin} + 2
  }
  \log(4d/\delta)
\end{align*}
where
\begin{align*}
  d_{\vP}
  & := \max_{i \in [d]} \sum_{j=1}^d \frac{P_{i,j}}{\pi_j}
  \leq
  \frac{1}{\pimin}
  \,.
\end{align*}

For simplicity, we assume that all of the blocks have length exactly $a$.
Under this assumption, 
the above bound (also with looser constants) gives that
with probability at least $1-\delta$:
\begin{equation}
\label{eq:indbound}
  \Norm{\frac1{\mu_H} \sum_{s=1}^{\mu_H} \wt\vZ_s}
  \leq
  \sqrt{
    \frac{12\log(4d/\delta)}{\mu_H\pimin}
  }
  +
  \frac{2\log(4d/\delta)}{\mu_H\pimin}
  \,.
\end{equation}

@@

It is not hard to see that $\beta(\bbP)$ in turn is upper bounded by 
the $(a-1)$-th beta mixing coefficient of the Markov chain $\{X_t\}$ (because the first
index in $H_{s+1}$ and the last index in $H_s$ are at a distance of $(a-1)$).
Hence, 
%of a signed measure $\mu$ over a measurable space 
%$(\Omega,\mathcal{A})$, defined by 
%$\Norm{\mu}_{\tv} = \frac12(\mu^+(\Omega)+\mu^-(\Omega))$, 
%where $\mu^+$ and $\mu^-$ are the positive and negative parts of measure $\mu$.
by a well-known result 
\[
  \beta(\bbP)
  \leq
  \frac{1}{\pimin} \exp(-(a-1)\gap)
\]
\citep[see, for instance,][]{bradley05}. \todoc{Which theorem? I did not find this result there. Can anyone please refine this citation?}
It follows that if the block length $a$ satisfies
\begin{equation}
  \label{eq:block-length2}
  a \geq 1 + \frac{1}{\gap}\log\frac{n}{\delta\pimin}
  ,
\end{equation}
then the failure probability increases from $\delta$ to at most
$2\delta$.
%Note that this condition implies \eqref{eq:block-length}.
